---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(picosatire)

if (FALSE) {
# Check for functions which do not have an @examples block for roxygen
system("grep -c examples man/*Rd", intern = TRUE) |> 
  grep(":0$", x = _, value = TRUE)
}

if (FALSE) {
  covr::report(covr::package_coverage(
    line_exclusions = list()
  ))
}

if (FALSE) {
  pkgdown::build_site(override = list(destination = "../coolbutuseless.github.io/package/picosatire"))
}

# Makevars options to do some deep testing for CRAN

# Type conversions are sane
# PKG_FLAG=-Wconversion

# Pointer overflow checks i.e. dodgy pointer arithmetic
# PKG_CFLAGS+=-fsanitize=pointer-overflow -fsanitize-trap=pointer-overflow
# Then run in the debugger:
# R -d lldb 
# run
# testthat::test_local()

```

# picosatire

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![CRAN](https://www.r-pkg.org/badges/version/picosatire)](https://CRAN.R-project.org/package=picosatire)
[![R-CMD-check](https://github.com/coolbutuseless/picosatire-dev/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/picosatire-dev/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{picosatire}` is a SAT solver for R using the [PicoSAT](https://fmv.jku.at/picosat/) 
library.

See the [`{satire}` package](https://github.com/coolbutuseless/satire) for a more 
comprehensive introduction to building SAT problems in R.


This packages includes PicoSAT release 965 (with API version 953).

### What's in the box

* `pico_solve_literals()` solve problems which are presented as a sequence 
  of literals (i.e. an integer vector)
* `pico_solve_satire()` solve problems developed as a `sat` object with the
  [`{satire}` package](https://github.com/coolbutuseless/satire)
* `pico_version()` returns version information about the PicoSAT C library
  included in this package.



## Installation

<!-- This package can be installed from CRAN -->

<!-- ``` r -->
<!-- install.packages('picosatire') -->
<!-- ``` -->

You can install the latest development version from
[GitHub](https://github.com/coolbutuseless/picosatire) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/picosatire')
```

Pre-built source/binary versions can also be installed from
[R-universe](https://r-universe.dev)

``` r
install.packages('picosatire', repos = c('https://coolbutuseless.r-universe.dev', 'https://cloud.r-project.org'))
```


### Example: `pico_solve_literals()`

Integer literals are a standard encoding for SAT solving.

Using literals involves developing Boolean expressions in conjunctive normal form.
Variables are then replaced by integers - the positive version of the integer
indicating `TRUE` and the negative version indicates `FALSE`.  Zeros are used
to represent logical `AND`.

When is the following expression true?

`(a | b) & (a | !c)`

Let:

* a = 1
* b = 2
* c = 3

So expresion becomes:  `(1 | 2) & (1 | -3)`.

Drop the logical `OR` statements, replace logical `AND` with `0`, and add a 
trailing zero on the end.  I.e.

`1 2 0 1 -3 0`

Asking PicoSAT to solve this expression:

```{r}
pico_solve_literals(c(1L, 2L, 0L, 1L, -3L, 0L), max_solutions = 8)
```

This indicates that there are 5 solutions to the problem.  The fifth solution
`-1 2 -3` says that when `a = FALSE, b = TRUE and c = FALSE` the original 
statement is `TRUE`


### Example: `pico_solve_satire()`

Using `{satire}` to define the problem makes creation easier and the solution
easier to interpret.

```{r}
library(satire)
sat <- sat_new()
sat_add_exprs(sat, "(a | b) & (a | !c)")
sat
pico_solve_satire(sat, max_solutions = 8)
```



### Example: Unsatisfiable problems

If the SAT problem is unsatisfiable, the returned solution will be `NULL`

E.g.

```{r}
library(satire)
sat <- sat_new()
sat_add_exprs(sat, "a & !a")
sat
pico_solve_satire(sat, max_solutions = 8)
```

















